\textbf{Authorship Statement}

\textit{Chapter 1: General Introduction.} I was the sole author for this first chapter. My co-promoter, Michiel de Vries (\textbf{MV}), helped through the editing process. My promoter, Fred van Eeuwijk (\textbf{FE}), assisted by reviewing latter drafts. Generative artificial intelligence was also used. Specifically, the \texttt{deepseek-chat \& deepseek-reasoner} model provided by DeepSeek (Version published June, 2025) was utilized for initial editing and review of early drafts.

\textit{Chapter 2: Little heterosis found in Diploid Hybrid Potato.} I was the primary author of this chapter with initial conceptualization done by me and \textbf{FE}. I conducted the research for this chapter with regular technical input from Chaozhi Zheng (\textbf{CZ}) and \textbf{FE}. \textbf{MV} assisted with content-related feedback.

\textit{Chapter 3: Efficient genomic prediction of yield and dry matter in hybrid potato.} I was the primary author for this research chapter with the idea for the research coming from myself and \textbf{FE}. \textbf{CZ} helped me in early design and technical discussions for the research. Research and the writing of the first draft ere conducted by me. \textbf{FE}, \textbf{MV}, \textbf{CZ} all assisted in the capacity of reviewers in the writing for this chapter.

\textit{Chapter 4: Digging up QTL: Evaluating the efficiency of marker-assisted selection in hybrid potato.} The initial concept and conducting of research was done by me. \textbf{FE} helped steer the research and determine which results were most relevant. \textbf{MV} helped with contextualizing the impact of this research.

\textit{Chapter 5: Evaluating multiallelism in genomic prediction applications.} The initial conceptualization for this chapter came from myself and \textbf{FE}. \textbf{CZ} provided technical assistance and supervision with the identity by descent estimation. Statistical modelling frameworks were selected by me. Generation of results and writing of the first draft was also done by me with \textbf{FE}, \textbf{MV}, \& \textbf{CZ} providing assistance through the editing process.

\textit{Chapter 6: General Discussion.} This last chapter was written by me. \textbf{MV} assisted in feedback over the narrative arc with helpful technical commentary provided by \textbf{FE} in later drafts.
